Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-12-17 demystify.tex
updates

\newcommand{\chapterinfo}[1]{%
  \phantomsection
  \addcontentsline{toc}{chapterinfo}{\color{black}{#1}}%
}
\newcommand{\sectioninfo}[1]{%
  \phantomsection
  \addcontentsline{toc}{sectioninfo}{\color{black}{#1}}%
}
\newcommand{\subsectioninfo}[1]{%
  \phantomsection
  \addcontentsline{toc}{subsectioninfo}{\color{black}{#1}}%
}

@@

\chapterinfo{\emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities in the US is not any harder than in other countries.}

@@

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Common concerns include (i) the \textbf{difficulty of getting admitted}, (ii) the \textbf{cost of graduate study} in the US, and for some, the fear of \textbf{discrimination} based on gender identity, sexual orientation, or cultural background.

Much of these uncertainties stem from the lack of information and guidance on the admission process in the US (social places like Facebook and Reddit are full of confusion and contradictory information).  This document aims to address these concerns.

@@

\paragraph{Discrimination is not tolerated} Third, higher education in the US take a very strong stance against discrimination, especially at the university level.
Many resources are available to encourage and support students from minority groups to pursue research, e.g., specific scholarships and fellowships. Faculty in CS are also \emph{very} familiar with and are highly encouraged to support such students in their group. There are also incentives, such as dedicated funding, to motivate faculty to recruit and mentor minority students (\autoref{sec:urm}).

@@

No, not in the US (other countries often do though~\autoref{sec:non-us-differences}).  In the US, it is common to apply for a PhD program directly after your undergrad.  Most CS PhD programs are designed so that students can get MS degree ``along the way'' to PhD, e.g., after finishing the 2-year course work.

@@

\begin{table}
\caption{Comparison of US and Other PhD Programs}\label{tab:us-vs-other}
\center
\small
\begin{tabular}{c|c|c}
\toprule
\textbf{Aspect} & \textbf{US PhD Programs} & \textbf{Other Countries} \\
\midrule
Duration & 5-7 years & 3-5 years \\
MS Required & Not required & Often required \\
Coursework Required & Yes (first 2 years) & No \\
Research Proposal Required & No & Yes (in some countries) \\
Academic (Faculty) Job & Direct & Postdoc \\
Work Life Balance & Less & More \\
\bottomrule
\end{tabular}
\end{table}

@@

\sectioninfo{Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members.}

@@

\section{Admission Committee (adcom)}\label{sec:adcom}

@@

\chapter{Letters of Recommendation (LoR)}\label{chap:lor}
\chapterinfo{\emph{LORs are very important}, but only if they are personalized and talk about your research ability.}

@@

\chapter{Research Experience}\label{chap:research-experience}
\chapterinfo{\emph{Publications can greatly help}. Papers in good venues are concrete evidence that you have successfully engaged in research. But they are not required!}

@@

\chapter{Statement of Purpose (SOP)}\label{chap:sop}
\chapterinfo{\emph{SOP is important}. Write it in such a way that makes you \emph{stand out}.}

@@

\chapter{Your School and Grades}\label{chap:your-school}
\chapterinfo{\emph{High grades probably won't help much} (unless you're from a very top and well-known school), but bad ones likely will raise concerns.}

@@

\chapter{Standard Tests (GRE and IELTS)}\label{chap:standard-tests}
\chapterinfo{\emph{Standard tests are not important}. GRE typically \emph{is not} required. For standard English tests (not required for domestic students), just do enough to pass the minimum requirements.}

@@

\part{After You Apply}\label{part:after-apply}
\chapter{Interview and the Waiting Game}\label{chap:interview}
\chapterinfo{Getting an interview is typically a \emph{good sign}; but no interview does not mean rejection}

@@

\section{Notification Timeline: Why rejection letters are sent so late?}\label{sec:late-rejection}
\sectioninfo{Grad programs often wait for the accepted students to make their decisions, typically by April 15, before sending out rejection letters.}

>>>>

\newcommand{\chapterinfo}[1]{%
  \phantomsection
  \addcontentsline{toc}{chapterinfo}{\color{black}{\emph{#1}}}%
}
\newcommand{\sectioninfo}[1]{%
  \phantomsection
  \addcontentsline{toc}{sectioninfo}{\color{black}{\emph{#1}}}%
}
\newcommand{\subsectioninfo}[1]{%
  \phantomsection
  \addcontentsline{toc}{subsectioninfo}{\color{black}{\texttt #1}}%
}

@@

\chapterinfo{Yes, definitely.  CS PhD study in the US is fully funded and admission into good universities in the US is not any harder than in other countries.}

@@

International students, especially those from less well-known universities and smaller countries, often wonder if they should apply to a CS PhD program in the US. Common concerns include (i) the \textbf{difficulty of getting admitted}, (ii) the \textbf{cost of graduate study} in the US, and for some, the fear of \textbf{discrimination} e.g., based on gender identity or cultural background.
Much of these uncertainties stem from the lack of information and guidance on the admission process in the US (social places like Facebook and Reddit are full of confusion and contradictory information).  This document aims to address these concerns.

@@

\paragraph{Discrimination is not tolerated} Third, not only is discrimination illegal, diversity and inclusion are highly valued in US universities. Many universities take great pride in their diverse student body and faculty.

In fact, many resources are available to encourage and support students from minority and diverse groups to pursue higher education and research, e.g., specific scholarships and fellowships. Faculty in CS are also \emph{very} familiar with and are highly encouraged to support such students in their group. There are also incentives, such as dedicated funding and awards, to motivate faculty to recruit and mentor minority students (\autoref{sec:urm}).

@@

No, while other countries often require an MS for PhD student in CS (~\autoref{sec:non-us-differences}), it is common in he US to apply for a PhD program directly after a 4-year undergrad program (e.g., after getting a B.S degree).  Most CS PhD programs are designed so that students can get MS degree ``along the way'' to PhD, e.g., after finishing the 2-year course work.

@@

\begin{table}
\caption{Comparison of the CS PhD program in the US and other countries}\label{tab:us-vs-other}
\center
\small
\begin{tabular}{c|c|c}
\toprule
\textbf{Aspect} & \textbf{US PhD Programs} & \textbf{Other Countries} \\
\midrule
Duration & 5-7 years & 3-5 years \\
MS Required & Not required & Often required \\
Coursework Required & Yes (first 2 years) & No \\
Research Proposal Required & No & Yes (in some countries) \\
Academic (Faculty) Job & Direct & Postdoc \\
Work Life Balance & Less & More \\
\bottomrule
\end{tabular}
\end{table}

@@

\chapterinfo{Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members.}

@@

\section{Admission Committee (adcom)}\label{sec:adcom}
\sectioninfo{Adcom members are faculty who evaluate your application.}

@@

\chapter{Letters of Recommendation (LoR)}\label{chap:lor}
\chapterinfo{LORs are very important, but only if they are personalized and talk about your research ability.}

@@

\chapter{Research Experience}\label{chap:research-experience}
\chapterinfo{Publications can greatly help. Papers in good venues are concrete evidence that you have successfully engaged in research. But they are not required!}

@@

\chapter{Statement of Purpose (SOP)}\label{chap:sop}
\chapterinfo{SOP is important. Write it in such a way that makes you \emph{stand out}.}

@@

\chapter{Your School and Grades}\label{chap:your-school}
\chapterinfo{High grades probably won't help much, but bad ones likely will raise concerns.}

@@

\chapter{Standard Tests (GRE and IELTS)}\label{chap:standard-tests}
\chapterinfo{Standard tests are not important. GRE typically \emph{is not} required. For standard English tests (not required for domestic students), just do enough to pass the minimum requirements.}

@@

\part{After You Apply}\label{part:after-apply}
\chapter{Interview and the Waiting Game}\label{chap:interview}
\chapterinfo{Getting an interview is typically a good sign; but no interview does not mean rejection}

@@

\section{Notification Timeline: Why rejection letters are sent so late?}\label{sec:late-rejection}
\sectioninfo{Grad programs often wait for accepted students to make their decisions, typically by April 15, before sending out rejection letters.}
